\Ace: Active Typechecking and Translation Inside a Python

### 2014-03-26 ace-oopsla14/ace-oopsla14.tex
OOPSLA14 final

%\documentclass[12pt]{article}
\documentclass[10pt,preprint]{sigplanconf}
\newcommand{\lamAce}{\lambda_{\text{Ace}}}
% The following \documentclass options may be useful:
%
% 10pt          To set in 10-point type instead of 9-point.
% 11pt          To set in 11-point type instead of 9-point.
% authoryear    To obtain author/year citation style instead of numeric.
\usepackage{listings}
\usepackage[usenames,dvipsnames]{color}
\usepackage{mathpartir}
\usepackage{amsmath}
\usepackage{amssymb} 
\usepackage{amsthm}
\usepackage{ stmaryrd }
\usepackage{verbatimbox}
\input{../att-icfp14/macros-atlam}
\usepackage{alltt}
\renewcommand{\ttdefault}{txtt}
\usepackage{color}
\usepackage[usenames,dvipsnames,svgnames,table]{xcolor}
\definecolor{mauve}{rgb}{0.58,0,0.82}
\definecolor{light-gray}{gray}{0.5}
\usepackage{listings}

    \makeatletter
%
% \btIfInRange{number}{range list}{TRUE}{FALSE}
%
% Test if int number <number> is element of a (comma separated) list of ranges
% (such as: {1,3-5,7,10-12,14}) and processes <TRUE> or <FALSE> respectively
%
        \newcount\bt@rangea
        \newcount\bt@rangeb

@@

\lstset{
  language=Python,
  showstringspaces=false,
  formfeed=\newpage,
  tabsize=4,
  commentstyle=\itshape\color{light-gray},
  basicstyle=\ttfamily\scriptsize,
  morekeywords={lambda, self, assert, as,type},
  numbers=left,
  numberstyle=\scriptsize\color{light-gray}\textsf,
  xleftmargin=2em,
  stringstyle=\color{mauve}
}
\lstdefinestyle{Bash}{
    language={}, 
    numbers=left,
    numberstyle=\scriptsize\color{light-gray}\textsf,
    moredelim=**[is][\color{blue}\bf\ttfamily]{`}{`},
}
\lstdefinestyle{OpenCL}{
	language=C++,
	morekeywords={kernel, __kernel, global, __global, size_t, get_global_id, sin, printf, int2}
}

@@

\title{\Ace: An Extensible Static Type System Inside a Python}

@@

We introduce Ace, an extensible statically typed language embedded within and compatible with Python, a widely-adopted dynamically typed language. Python serves as Ace's type-level language. Rather than building in a particular set of type constructors, Ace introduces a novel extension mechanism, \emph{active typechecking and translation}, organized around a bidirectional type system that inverts control over typechecking and translation to user-defined type constructors according to a protocol that cannot cause ambiguities and type safety issues at link-time. In addition to describing a full-scale language design, we give a simplified calculus that describes the foundations of this mechanism and show that, as implemented in Ace, it is flexible enough to admit practical library-based implementations of types drawn from functional, object-oriented, parallel and domain-specific languages. We also show how to construct safe, natural and extensible foreign function interfaces using a novel form of ``staged'' type propagation from Python into Ace and then through to OpenCL. % can be used to enable type system extensions and demonstrates how type safety can be maintained by lifting a technique from the typed compilation literature into the language.%To show that the method scales, we give a full implementation of the OpenCL programming language as a library. %We briefly examine the type theoretic foundations of this approach and give modular criteria that guarantee that an extension is safe and modular. %Ace can be used from the shell or interactively from within Python\todo{take this sentence out?}.
%Researchers developing languages and abstractions for high-performance computing must consider a number of design criteria, including performance, verifiability, portability and ease-of-use. Despite the deficiencies of legacy tools and the availability of seemingly superior options, end-users have been reluctant to adopt new language-based abstractions. We argue that this can be largely attributed to a failure to consider three additional criteria: continuity, extensibility\- and interoperability. This paper introduces Ace, a language that aims to satisfy this more comprehensive set of design criteria. To do so, Ace introduces several novel compile-time mechanisms, makes principled design choices, and builds upon existing standards in HPC, particularly Python and OpenCL. OpenCL support, rather than being built into the language, is implemented atop an extensibility mechanism that also admits abstractions drawn from other seemingly disparate paradigms. The core innovation underlying this and other features of Ace is a novel reification of types as first-class objects at compile-time, representing a refinement to the concept of active libraries that we call \emph{active types}. We validate our overall design by considering a case study of a simulation framework enabling the modular specification and efficient execution of ensembles of neural simulations across a cluster of GPUs.
\end{abstract}

@@

%Ace can be used from the shell, producing source files that can be further compiled or executed by external means. Ace functions targeting a language with Python bindings (e.g. Python itself, as well as the others just described) can also be compiled and invoked interactively from Python. 
Ace supports a form of staged \emph{ad hoc} polymorphism (based on singleton types) wherein the ``dynamic type'' of a Python value can determine a static type and trigger active typechecking and translation ``just-in-time'' when Python code calls a generic Ace function. We show how this can significantly streamline interactive workflows (e.g. in scientific computing) that use Python for orchestration and exploratory tasks (e.g. plotting) but call into a statically-typed  language via a foreign-function interface (FFI) for performance- and safety-critical portions. We provide a complete implementation of the OpenCL type system (a language for low-level data parallel programming on GPUs), show several higher-level extensions to it, and discuss bindings that bridge between the type systems of Ace, OpenCL and Python's standard numerics package, \verb|numpy|, all as a library (contribution 3, Sec. \ref{targets}).%We will discuss in detail as our primary realistic case study.

@@

In Sec.  \ref{related}, we place Ace in the context of related work on language extensibility, term rewriting, staged  compilation, bidirectional typechecking, type-level computation and typed compilation. We conclude in Sec. \ref{discussion} by summarizing the key features needed by a host language to support active typechecking and translation, and discussing present limitations and potential future work. The appendix has details omitted for concision as well as several more examples, including functional datatypes, flexible complex numbers and a previously little-explored type system for tracking string invariants using regular expressions statically (which we show being used in Listing \ref{example}, amongst  other examples).

@@

Types are constructed programmatically during execution of the compilation script. This stands in contrast to many  contemporary statically-typed languages, where types (e.g. datatypes, classes, structs) can only be declared. Put another way, Ace supports \emph{type-level computation} and Python is its type-level language (discussed further in Sec. \ref{theory} and Sec. \ref{related}). 
%The compilation script can assign shorter names to types for convenience (removing the need for  facilities like \verb|typedef| in C or \verb|type| in Haskell). 
In our example, we see several types being constructed:
\begin{enumerate}
\item On line 9, we construct a functional record type with a single (immutable) field named \verb|amount| with type \verb|decimal[2]|, which classifies decimal numbers with two decimal places. \verb|record| and \verb|decimal| are \emph{indexed type constructors}. We will provide more details in Sec. \ref{att}, but note that syntactically, \verb|record| overloads subscripting and borrows Python's syntax for array slices (e.g. \verb|a[min:max]|) to approximate conventional functional notation for type annotations. Note also that the field name could equivalently have been written \lstinline{'a' + 'mount'}, again emphasizing that types and indices are constructed programatically by a Python script.
\item On lines 10-13, we construct another record type. The field \verb|name| has type \verb|string|, defined in \verb|examples.py|, while the fields \verb|account_num| and \verb|routing_num| have more interesting types, classifying strings guaranteed statically to be in a regular language specified statically by a  regular expression pattern (written, to avoid needing to escape backslashes, using Python's \emph{raw string literals}). The appendix specifies the typechecking rules, based on \cite{fulton-thesis}. To our knowledge, this type system has not previously been implemented. We include it to emphasize that we aim to support specialized type systems,  not just general purpose constructs like records, numbers and strings.
\item 
%Prototypic inheritance  is a feature of some dynamically-typed languages (most notably, Self \cite{Ungar:Smith:oopsla:1987} and Javascript). 
On line 14, we construct a simple \emph{prototypic object type} \cite{Lie86}. The type \verb|Transfer| classifies terms consisting of a \todo{what is the right term for this?}\emph{fore} of type \verb|A| and a \emph{prototype} of type \verb|C|. If a field cannot be found in the fore, the type system will delegate (here, statically) to the prototype. This makes it easy to share the values of the fields of a common prototype amongst many fores, here to allow us to describe multiple transfers differing only in amount.
\end{enumerate}
\subsection{Typed Functions}

@@

On lines 16-21, we see the function \verb|log_transfer|. We follow Python conventions by starting with a documentation string  that informally specifies the behavior of the function. Before moving into the body, however, we  also write a \emph{type signature} (line 17) stating that the type of \verb|t| is \verb|Transfer|, the prototypic object type described above. As a result, we can assume on line 19 that \verb|t.amount| and \verb|t.name| have types \verb|decimal[2]| and \verb|string|, respectively. We will return to the details of \verb|print| and the form \verb|[string](t.amount)|, which performs an explicit conversion, shortly.\todo{when?}

@@

def print_transfer(t):
  print "Transferring %s to %s." % (
    __ace_0.decimal_to_str(t[0],2), t[1][0])

@@

To typecheck, translate and execute the code in Listing \ref{example}, we have two choices: do so externally at the shell, or perform compilation interactively (or implicitly) from within Python. We will begin with the former, then return to interactive and implicit compilation in Sec. \ref{targets}.% with our OpenCL examples.

@@

The Ace compiler is itself a Python library, and \verb|acec| is a simple Python script that invokes it, operating in two steps:
\begin{enumerate}
\item It evaluates the compilation script to completion.
\item For any top-level bindings that are Ace functions in the final environment (instances of \verb|ace.TypedFn|, as we will discuss), it initiates active type-checking and translation (Sec. \ref{att}). If no type errors are discovered, the translations are collected (obeying order dependencies) and emitted, with file extension(s) determined by the target(s) in use, discussed in Sec. \ref{targets}. Here our target is  the default target associated with the base \verb|py|, which emits Python 2.6 files. If a type error is discovered, no file is emitted and the error is displayed on the console.
\end{enumerate}

@@

Enabling the addition of new forms to a language in an open tool ecosystem is difficult. Indeed, it is the canonical example of the long-studied \emph{expression problem} \cite{wadler1998expression}. Although a number of approaches have been developed, we argue that many of them are overly permissive, making it difficult to reason compositionally about metatheoretic issues (e.g. type safety) and avoid ambiguities when extensions are combined (see Sec. \ref{related}). %They are also often quite complex. %Requiring that abstraction clients have the expertise to resolve these ambiguities manually is undesirable, and heading them off results in an unbounded, combinatoric explosion of situations to consider.
Ace's term forms, as we have seen, are fixed by Python's grammar, so Ace sidesteps the expression problem almost entirely (and thus inherits broad tool support, as described earlier). The key insight is that leaving the forms fixed does not mean the semantics must also be fixed. Instead of taking a syntax-directed view of extensibility, we take a type-directed view: users cannot add new forms but they can add new type constructors, which we empower with  more control over the semantics of existing forms (e.g. literals, as discussed above, but also nearly every other form, as we will now discuss) in a controlled manner.  %In Ace (but not in the theory in Sec. \ref{theory}), users can also repurpose existing forms flexibly within a delimited scope. 
The key features that characterize active typechecking and translation are: \begin{itemize}
\item a \emph{dispatch protocol} that delegates responsibility for typechecking and translating a term to:
  \begin{itemize}
  \item a type or type constructor extracted from a subterm
  \item a per-function base semantics (or simply \emph{base}) for forms for which this is not possible  (variables, literals without ascriptions and most statements)
  \end{itemize}
\item a mechanism for allowing the user to define new types, type constructors and bases from within the language (in particular, \emph{from within the type-level language}) rather than fixing them ahead of time
\end{itemize}
\begin{codelisting}
\begin{lstlisting}
from listing1 import py, A
@py
def example(a):
  {a : A}
  x = a.amount
  x = x + x
  return {
    y: {amount: x} (A),
    remark: "Creating an anonymous record"} (record)
\end{lstlisting}
\caption{[\texttt{listing6.py}] The example detailed in Sec. \ref{att}.}
\label{att-example}
\end{codelisting}

@@

\subsection{Base Decorators}
Decorators in Python (line 2) are syntactic sugar: we could equivalently have omitted line 3 and inserted the top-level statement \verb|example = py(example)| on the line after the function definition. Note, however, that \verb|py| is not a function but an instance of a class, \verb|examples.py.PyBase|, that inherits from \verb|ace.Base|. It can be used as a decorator because \verb|ace.Base| overloads the call operator, shown in Listing \ref{base}. The \verb|_process| function (not shown) \todo{show in appendix?}operates as follows:
\begin{enumerate}
\item The Python standard library is used to extract an abstract syntax tree (\verb|AST|) from the function.
\item The closure of the function, together with the globals in the module it is defined in, are extracted to reify its static environment.
\item The argument annotations are processed. If provided in the body, as in our example, \verb|_process| checks that the argument names are spelled correctly and evaluates the type-level expressions (e.g. \verb|A| above) in the static environment to produce a mapping of argument names to types, called the argument signature. If Python 3's annotations were used, this is not necessary.
%\item Whether the function is generic or not is determined.  We defer discussion of generic functions until Sec. \ref{targets}.
\end{enumerate}
By the end of Listing \ref{att-example}, \verb|example| is thus an instance of \verb|ace.TypedFn|. It could  have been constructed directly by calling the constructor on line 2 of Listing \ref{base} (this would be inconvenient, but could be useful for metaprogramming).

\subsection{Checking Typed Functions}
When the Ace compiler begins typechecking a \verb|TypedFn| (when asked to by \verb|acec|, or when it is  invoked interactively as we will show in Sec. \ref{targets}), it proceeds as follows:
\begin{enumerate}
\item An \verb|ace.Context| is constructed with a reference to the function (Listing \ref{context}, lines 2-3).
\item The base is asked to initialize the context. This can be seen on lines 2-4 of Listing \ref{pybase}, where the \verb|PyBase| base adds an attribute tracking local bindings (initially only the arguments) and the return type, which will be synthesized but initially is not known. \todo{briefly mention return type annotations?}
\item For each statement in the body (after the documentation string and type annotation, if present), the compiler calls the context's \verb|check| method, which in turn defers to a method of the base, \verb|check_|$F$, where $F$ is the form of the statement (taken directly from Python's \verb|ast| package \cite{python}). These methods are responsible for checking that the statement is well-typed (raising an \verb|ace.TypeError| if not) and having any necessary effect on the context. 
\item The base synthesizes a type for the function as a whole via the \verb|syn_FunctionDef_outer| method. Here, an \verb|arrow| type is synthesized (lines 26-29) based on the argument signature and the synthesized return type.
\end{enumerate}

Because bases are user-defined objects, this represents the first instance of the compiler delegating control to a library.
\subsection{Bidirectional Typechecking}
\begin{codelisting}
\begin{lstlisting}
class TypedFn(object):
  def __init__(self, base, ast, static_env, arg_sig):
    # ...
  # ...
class Base(object):
  def __call__(self, f):
    (ast, static_env, arg_sig) = _process(f)
    return TypedFn(self, ast, static_env, arg_sig)
  # ...
\end{lstlisting}
\caption{A portion of the \texttt{ace} core showing how a base can be used as a decorator to construct a typed function.}
\label{base}
\end{codelisting}

@@

  def check_Assign(self, ctx, s):
    ace.util.unsupported_if(len(s.targets) != 1
      or not isinstance(s.targets[0], ast.Name))
    x, e = s.targets[0].id, s.value # see ast module
    if x in ctx.locals:
      ctx.ana(e, ctx.locals[x])
    else:
      ty = ctx.syn(e)
      ctx.locals[x] = ty

  def trans_Assign(self, ctx, target, s):
    return target.direct_translation(ctx, s)

@@

  def trans_Name(self, ctx, target, e):
    if e.id in ctx.locals:
      return target.direct_translation(ctx, e)
    else: # (Sec. 5)
      return self.trans_lifted(ctx, target, e)
  # ...
  default_target = PyTarget()   
  default_Dict_asc = dyn
  # ...
py = PyBase()
\end{lstlisting}
\caption{A portion of the base used in our examples thus far, defined in the \texttt{examples.py} package.}
\label{pybase}
\end{codelisting}

@@

  def check(self, s):
    if isinstance(s, ast.Assign):
      self.fn.base.check_Assign(self, s)
    # ... other statement forms similar

  def syn(self, e):
    if instanceof(e, ast.Name):
      delegate = self.fn.base
      ty = delegate.syn_Name(self, e)
    elif instanceof(e, ast.Attribute):
      delegate = self.syn(e.value)
      ty = delegate.syn_Attribute(self, e)
    # ... other compound forms similar (cf appendix)
    elif isinstance(e, ast.Dict):
      return self.ana(self, e, 
        self.fn.base.default_Dict_asc)
    # ... other unascribed literal forms similar
    elif is_ascribed_Dict(e):
      lit, delegate = get_lit(e)
      if issubclass(delegate, Type): # tycon
        ty = delegate.syn_Dict(self, lit)
      else:
        return self.ana_Dict(lit, delegate)
    # ... other ascribed literal forms similar
    e.delegate = delegate
    e.ty = ty
    return ty

@@

  def __eq__(self, other): 
    if self.anon or other.anon: 
      return self is other
    else: 
      return (isinstance(other, record) 
        and self.sig.equiv(other.sig))

  def ana_Dict(self, ctx, e):
    for f, v in zip(e.keys, e.values):
      if f.id in self.sig:
        ctx.ana(v, self.sig[f.id])
      else:
        raise ace.TypeError("...bad field...", f)
    if len(self.sig) != len(e.keys):
      raise ace.TypeError("...missing field...", e)

  @classmethod # (when ascription is a type constructor)
  def syn_Dict(cls, ctx, e):
    sig = Sig()
    for f, v in zip(e.keys, e.values):
      sig[f] = ctx.syn_ty(v)
    return cls(sig, anon=True)

@@

The method \verb|check_Assign| on lines \ref{pybase}.6-\ref{pybase}.14 is called. Because the locals dictionary in \verb|ctx| does not yet know what type \verb|x| should be, the base asks to synthesize a type for the expression being bound, \verb|a.amount|, by calling \verb|ctx.syn|.

>>>>

%\documentclass[12pt]{article}
\documentclass[10pt,preprint]{sigplanconf}
\newcommand{\lamAce}{\lambda_{\text{Ace}}}
% The following \documentclass options may be useful:
%
% 10pt          To set in 10-point type instead of 9-point.
% 11pt          To set in 11-point type instead of 9-point.
% authoryear    To obtain author/year citation style instead of numeric.
\usepackage{listings}
\usepackage[usenames,dvipsnames]{color}
\usepackage{mathpartir}
\usepackage{amsmath}
\usepackage{amssymb} 
\usepackage{amsthm}
\usepackage{ stmaryrd }
\usepackage{verbatimbox}
\input{../att-icfp14/macros-atlam}
\usepackage{alltt}
\renewcommand{\ttdefault}{txtt}
\usepackage{color}
\usepackage[usenames,dvipsnames,svgnames,table]{xcolor}
\definecolor{mauve}{rgb}{0.58,0,0.82}
\definecolor{light-gray}{gray}{0.5}
\usepackage{listings}
\usepackage{wasysym}
    \makeatletter
%
% \btIfInRange{number}{range list}{TRUE}{FALSE}
%
% Test if int number <number> is element of a (comma separated) list of ranges
% (such as: {1,3-5,7,10-12,14}) and processes <TRUE> or <FALSE> respectively
%
        \newcount\bt@rangea
        \newcount\bt@rangeb

@@

\lstset{
  language=Python,
  showstringspaces=false,
  formfeed=\newpage,
  tabsize=4,
  commentstyle=\itshape\color{light-gray},
  basicstyle=\ttfamily\scriptsize,
  morekeywords={lambda, self, assert, as, cls},
  numbers=left,
  numberstyle=\scriptsize\color{light-gray}\textsf,
  xleftmargin=2em,
  stringstyle=\color{mauve}
}
\lstdefinestyle{Bash}{
    language={}, 
    numbers=left,
    numberstyle=\scriptsize\color{light-gray}\textsf,
    moredelim=**[is][\color{blue}\bf\ttfamily]{`}{`},
}
\lstdefinestyle{OpenCL}{
	language=C++,
	morekeywords={kernel, __kernel, global, __global, size_t, get_global_id, sin, printf, int2}
}

@@

\title{\Ace: Active Typechecking and Translation Inside a Python}

@@

We introduce Ace, an extensible statically typed language embedded within and compatible with Python, a widely-adopted dynamically typed language. Python serves as Ace's type-level language. Rather than building in a particular set of type constructors, Ace introduces a novel extension mechanism, \emph{active typechecking and translation}, organized around a bidirectional type system that inverts control over typechecking and translation to user-defined type constructors according to a protocol that cannot cause ambiguities and type safety issues at link-time. In addition to describing a full-scale language design, we give a simplified calculus that describes the foundations of this mechanism and show that, as implemented in Ace, it is flexible enough to admit practical library-based implementations of types drawn from functional, object-oriented, parallel and domain-specific languages. 
%We also show how to construct safe, natural and extensible foreign function interfaces using a novel form of ``staged'' type propagation from Python into Ace and then through to OpenCL. % can be used to enable type system extensions and demonstrates how type safety can be maintained by lifting a technique from the typed compilation literature into the language.%To show that the method scales, we give a full implementation of the OpenCL programming language as a library. %We briefly examine the type theoretic foundations of this approach and give modular criteria that guarantee that an extension is safe and modular. %Ace can be used from the shell or interactively from within Python\todo{take this sentence out?}.
%Researchers developing languages and abstractions for high-performance computing must consider a number of design criteria, including performance, verifiability, portability and ease-of-use. Despite the deficiencies of legacy tools and the availability of seemingly superior options, end-users have been reluctant to adopt new language-based abstractions. We argue that this can be largely attributed to a failure to consider three additional criteria: continuity, extensibility\- and interoperability. This paper introduces Ace, a language that aims to satisfy this more comprehensive set of design criteria. To do so, Ace introduces several novel compile-time mechanisms, makes principled design choices, and builds upon existing standards in HPC, particularly Python and OpenCL. OpenCL support, rather than being built into the language, is implemented atop an extensibility mechanism that also admits abstractions drawn from other seemingly disparate paradigms. The core innovation underlying this and other features of Ace is a novel reification of types as first-class objects at compile-time, representing a refinement to the concept of active libraries that we call \emph{active types}. We validate our overall design by considering a case study of a simulation framework enabling the modular specification and efficient execution of ensembles of neural simulations across a cluster of GPUs.
\end{abstract}

@@

%Ace can be used from the shell, producing source files that can be further compiled or executed by external means. Ace functions targeting a language with Python bindings (e.g. Python itself, as well as the others just described) can also be compiled and invoked interactively from Python. 
%Ace supports a form of staged \emph{ad hoc} polymorphism (based on singleton types) wherein the ``dynamic type'' of a Python value can determine a static type and trigger active typechecking and translation ``just-in-time'' when Python code calls a generic Ace function. We show how this can significantly streamline interactive workflows (e.g. in scientific computing) that use Python for orchestration and exploratory tasks (e.g. plotting) but call into a statically-typed  language via a foreign-function interface (FFI) for performance- and safety-critical portions. We provide a complete implementation of the OpenCL type system (a language for low-level data parallel programming on GPUs), show several higher-level extensions to it, and discuss bindings that bridge between the type systems of Ace, OpenCL and Python's standard numerics package, \verb|numpy|, all as a library (contribution 3, Sec. \ref{targets}).%We will discuss in detail as our primary realistic case study.

@@

In Sec.  \ref{related}, we place Ace in the context of related work on language extensibility, term rewriting, staged  compilation, bidirectional typechecking, type-level computation and typed compilation. We conclude in Sec. \ref{discussion} by summarizing the key features needed by a host language to support active typechecking and translation, and discussing present limitations and potential future work. %The appendix has details omitted for concision as well as several more examples, including functional datatypes, flexible complex numbers and a previously little-explored type system for tracking string invariants using regular expressions statically (which we show being used in Listing \ref{example}, amongst  other examples).

@@

Types are constructed programmatically during execution of the compilation script. This stands in contrast to many  contemporary statically-typed languages, where types (e.g. datatypes, classes, structs) can only be declared. Put another way, Ace supports \emph{type-level computation} and Python is its type-level language (discussed further in Sec. \ref{theory} and Sec. \ref{related}). 
%The compilation script can assign shorter names to types for convenience (removing the need for  facilities like \verb|typedef| in C or \verb|type| in Haskell). 
In our example, we see several types being constructed:
\begin{enumerate}
\item On line 9, we construct a functional record type with a single (immutable) field named \verb|amount| with type \verb|decimal[2]|, which classifies decimal numbers with two decimal places. \verb|record| and \verb|decimal| are \emph{indexed type constructors}. We will provide more details in Sec. \ref{att}, but note that syntactically, \verb|record| overloads subscripting and borrows Python's syntax for array slices (e.g. \verb|a[min:max]|) to approximate conventional functional notation for type annotations. Note also that the field name could equivalently have been written \lstinline{'a' + 'mount'}, again emphasizing that types and indices are constructed programatically by a Python script.
\item On lines 10-13, we construct another record type. The field \verb|name| has type \verb|string|, defined in \verb|examples.py|, while the fields \verb|account_num| and \verb|routing_num| have more interesting types, classifying strings guaranteed statically to be in a regular language specified statically by a  regular expression pattern (written, to avoid needing to escape backslashes, using Python's \emph{raw string literals}). The appendix will give the typechecking rules, based on \cite{fulton-thesis}. To our knowledge, this type system has not previously been implemented. We include it to emphasize that we aim to support specialized type systems,  not just general purpose constructs like records and strings.
\item 
%Prototypic inheritance  is a feature of some dynamically-typed languages (most notably, Self \cite{Ungar:Smith:oopsla:1987} and Javascript). 
On line 14, we construct a simple \emph{prototypic object type} \cite{Lie86}. The type \verb|Transfer| classifies terms consisting of a \emph{fore} of type \verb|A| and a \emph{prototype} of type \verb|C|. If a field cannot be found in the fore, the type system will delegate (here, statically) to the prototype. This makes it easy to share the values of the fields of a common prototype amongst many fores, here to allow us to describe multiple transfers differing only in amount.
\end{enumerate}
\subsection{Typed Functions}

@@

On lines 16-21, we see the function \verb|log_transfer|. We follow Python conventions by starting with a documentation string  that informally specifies the behavior of the function. Before moving into the body, however, we  also write a \emph{type signature} (line 17) stating that the type of \verb|t| is \verb|Transfer|, the prototypic object type described above. As a result, we can assume on line 19 that \verb|t.amount| and \verb|t.name| have types \verb|decimal[2]| and \verb|string|, respectively. We will return to the details of \verb|print| and the form \verb|[string](t.amount)|, which performs an explicit conversion, in Sec. \ref{att}.

@@

def print_transfer(t):
  print ("Transferring %s to %s." % (
    __ace_0.decimal_to_str(t[0],2), t[1][0]))

@@

To typecheck, translate and execute the code in Listing \ref{example}, we have two choices: do so externally at the shell, or perform compilation interactively (or implicitly) from within Python. We will begin with the former. Interactive and implicit compilation are implemented but we do lack space to detail it in the present paper, choosing here to focus on the core mechanism.% with our OpenCL examples.

@@

The Ace compiler is itself a Python library, and \verb|acec| is a simple Python script that invokes it, operating in two steps:
\begin{enumerate}
\item It evaluates the compilation script to completion.
\item For any top-level bindings that are Ace functions in the final environment (instances of \verb|ace.TypedFn|, as we will discuss), it initiates active type-checking and translation (Sec. \ref{att}). If no type errors are discovered, the translations are collected (obeying order dependencies) and emitted, with file extension(s) determined by the target(s) in use, discussed further in Sec. \ref{att}. Here our target is  the default target associated with the base \verb|py|, which emits Python 2.6 files. If a type error is discovered, no file is emitted and the error is displayed on the console.
\end{enumerate}

@@

Enabling the addition of new forms to a language in an open tool ecosystem is difficult. Indeed, it is the canonical example of the long-studied \emph{expression problem} \cite{wadler1998expression}. Although a number of approaches have been developed, we argue that many of them are overly permissive, making it difficult to reason compositionally about metatheoretic issues (e.g. type safety) and avoid ambiguities when extensions are combined (see Sec. \ref{related}). %They are also often quite complex. %Requiring that abstraction clients have the expertise to resolve these ambiguities manually is undesirable, and heading them off results in an unbounded, combinatoric explosion of situations to consider.
Ace's term forms, as we have seen, are fixed by Python's grammar, so Ace sidesteps the expression problem almost entirely (and thus inherits broad tool support, as described earlier). The key insight is that leaving the forms fixed does not mean the semantics must also be fixed. Instead of taking a syntax-directed view of extensibility, we take a type-directed view: users cannot add new forms but they can add new type constructors, which we empower with  more control over the semantics of existing forms (e.g. literals, as discussed above, but also nearly every other form, as we will now discuss) in a controlled manner.  %In Ace (but not in the theory in Sec. \ref{theory}), users can also repurpose existing forms flexibly within a delimited scope. 
The key features that characterize active typechecking and translation are: \begin{itemize}
\item a \emph{dispatch protocol} that delegates responsibility for typechecking and translating a term to:
  \begin{itemize}
  \item a type or type constructor extracted from a subterm
  \item a per-function base semantics (or simply \emph{base}) for forms for which this is not possible  (variables, literals without ascriptions and most statements)
  \end{itemize}
\item a mechanism for allowing the user to define new types, type constructors and bases from within the language (in particular, \emph{from within the type-level language}) rather than fixing them ahead of time
\end{itemize}
\begin{codelisting}
\begin{lstlisting}
from listing1 import py, A, record
@py
def example(a):
  {a : A}
  x = a.amount
  x = x + x
  return {
    y: {amount: x} (A),
    remark: "Creating an anonymous record"} (record)
\end{lstlisting}
\caption{[\texttt{listing6.py}] The example detailed in Sec. \ref{att}.}
\label{att-example}
\end{codelisting}

@@

\subsection{Base Decorators}
Decorators in Python (line 2) are syntactic sugar: we could equivalently have omitted line 3 and inserted the top-level statement \verb|example = py(example)| on the line after the function definition. Note, however, that \verb|py| is not a function but an instance of a class, \verb|examples.py.PyBase|, that inherits from \verb|ace.Base|. It can be used as a decorator because \verb|ace.Base| overloads the call operator, shown in Listing \ref{base}. The \verb|_process| function (not shown) operates as follows:
\begin{enumerate}
\item The Python standard library is used to extract an abstract syntax tree (\verb|AST|) from the function.
\item The closure of the function, together with the globals in the module it is defined in, are extracted to reify its static environment.
\item The argument annotations are processed. If provided in the body, as in our example, \verb|_process| checks that the argument names are spelled correctly and evaluates the type-level expressions (e.g. \verb|A| above) in the static environment to produce a mapping of argument names to types, called the argument signature. If Python 3's annotations were used, this is not necessary.
%\item Whether the function is generic or not is determined.  We defer discussion of generic functions until Sec. \ref{targets}.
\end{enumerate}
By the end of Listing \ref{att-example}, \verb|example| is thus an instance of \verb|ace.TypedFn|. It could  have been constructed directly by calling the constructor on line 2 of Listing \ref{base} (this would be inconvenient, but could be useful for metaprogramming).

\subsection{Checking Typed Functions}
When the Ace compiler begins typechecking a \verb|TypedFn| (when asked to by \verb|acec|, or when it is  invoked interactively, not discussed in this paper), it proceeds as follows:
\begin{enumerate}
\item An \verb|ace.Context| is constructed with a reference to the function (Listing \ref{context}, lines 2-3).
\item The base is asked to initialize the context. This can be seen on lines 2-4 of Listing \ref{pybase}, where the \verb|PyBase| base adds an attribute tracking local bindings (initially only the arguments) and the return type, which will be synthesized but initially is not known. 
\item For each statement in the body (after the documentation string and type annotation), the compiler delegates to either the base, or of a type synthesized from a subterm, by calling a method named \verb|check_|$F$, where $F$ is the form of the statement (derived directly from Python's \verb|ast| package \cite{python}). These methods are responsible for checking that the statement is well-typed (raising an \verb|ace.TypeError| if not) and having any needed effect on the context. 
\item The base synthesizes a type for the function as a whole via the \verb|syn_FunctionDef_outer| method. Here, an \verb|arrow| type is synthesized (lines 26-29) based on the argument signature and the synthesized return type.
\end{enumerate}
The assignment statements on lines \ref{att-example}.5\footnote{In this section, we will need to refer to code in Listings \ref{att-example}, \ref{pybase}, \ref{context} and \ref{record}, so we adopt this syntactic convention to refer to line numbers for concision.} and \ref{att-example}.6 are checked by the base method \verb|check_Assign_Name| shown on lines \ref{pybase}.6-\ref{pybase}.14 and the return statement on line \ref{att-example}.7 is checked by the base's \verb|check_Return| on lines \ref{pybase}.19-\ref{pybase}.23. Both work similarly: if this is the first assignment to a particular name, or the first return statement seen, then the value must be able to synthesize a type in the current context. Otherwise, it is analyzed against the previously synthesized type.

\subsection{Bidirectional Typechecking}
\begin{codelisting}
\begin{lstlisting}
class TypedFn(object):
  def __init__(self, base, ast, static_env, arg_sig):
    # ...
  # ...
class Base(object):
  def __call__(self, f):
    (ast, static_env, arg_sig) = _process(f)
    return TypedFn(self, ast, static_env, arg_sig)
  # ...
\end{lstlisting}
\caption{A portion of the \texttt{ace} core showing how a base can be used as a decorator to construct a typed function.}
\label{base}
\end{codelisting}

@@

  def check_Assign_Name(self, ctx, s):
    x, e = s.target.id, s.value
    if x in ctx.locals:
      ctx.ana(e, ctx.locals[x])
    else:
      ty = ctx.syn(e)
      ctx.locals[x] = ty

  def trans_Assign_Name(self, ctx, target, s):
    return target.direct_translation(ctx, s)

@@

  def trans_Name(self, ctx, target, e):
    if e.id in ctx.locals:
      return target.direct_translation(ctx, e)
    else: # (Sec. 5)
      return self.trans_lifted(ctx, target, e)

  default_Dict_asc = default_Str_asc = dyn
  # ...
  def init_target(ctx): return PyTarget()   
py = PyBase()
\end{lstlisting}
\caption{A portion of the base used in our examples thus far, defined in the \texttt{examples.py} package.}
\label{pybase}
\end{codelisting}

@@

  def syn(self, e):
    if instanceof(e, ast.Name):
      delegate = self.fn.base
      ty = delegate.syn_Name(self, e)
    elif instanceof(e, ast.Attribute):
      delegate = self.syn(e.value)
      ty = delegate.syn_Attribute(self, e)
    # ... other compound forms similar (cf appendix)
    elif isinstance(e, ast.Str):
      return self.ana(self, e, 
        self.fn.base.default_Str_asc)
    # ... other unascribed literal forms similar
    elif is_ascribed_Dict(e):
      lit, delegate = get_lit(e)
      if issubclass(delegate, Type): # tycon
        ty = delegate.syn_Dict(self, lit)
      else:
        return self.ana_Dict(lit, delegate)
    # ... other ascribed literal forms similar
    e.delegate = delegate
    e.ty = ty
    return ty

@@

  @classmethod 
  def syn_Dict(cls, ctx, e):
    sig = Sig((f, ctx.syn_ty(v)) 
      for f, v in zip(e.keys, e.values))
    return cls(sig, anon=True)

  def ana_Dict(self, ctx, e):
    for f, v in zip(e.keys, e.values):
      if f.id in self.sig:
        ctx.ana(v, self.sig[f.id])
      else:
        raise ace.TypeError("...extra field...", f)
    if len(self.sig) != len(e.keys):
      raise ace.TypeError("...missing field...", e)

@@

%\paragraph{Active Type Synthesis} 
This method is defined on lines \ref{context}.5-\ref{context}.26. The relevant case in this method is the one on line \ref{context}.9, because \verb|a.amount| is of the form \verb|Attribute| according to Python's grammar. The context delegates to the type recursively synthesized for its value, \verb|a|. We recurse back into \verb|syn|, now taking the first branch for terms of the form \verb|ast.Name|. Synthesizing a type for a name is delegated to the base by calling its \verb|syn_Name| method. We can see its implementation for the base we are using on lines \ref{pybase}.37-\ref{pybase}.44. The identifier \verb|a| is an argument to the function, which the base included in the initial locals dictionary, so we hit a base case and the type \verb|A| is synthesized.
